The MRChem multiresolution analysis code for molecular electronic structure calculations: performance and scaling properties

### 2022-11-06 ReviewersReply.tex

% environment derived from framed.sty: see leftbar environment definition
\definecolor{formalshade}{rgb}{0.95,0.95,1}

\newcommand{\ajwt}[1]{{\color{red}{#1}}}
\newenvironment{formal}{%
  \def\FrameCommand{%
    \hspace{1pt}%
    {\color{darkblue}\vrule width 2pt}%
    {\color{formalshade}\vrule width 4pt}%
    \colorbox{formalshade}%
  }%
  \MakeFramed{\advance\hsize-\width\FrameRestore}%
  \noindent\hspace{-4.55pt}% disable indenting first paragraph
  \begin{adjustwidth}{}{7pt}%
  \vspace{2pt}\vspace{2pt}%
}
{%
  \vspace{2pt}\end{adjustwidth}\endMakeFramed%
}
%\addtolength{\topmargin}{-.6in}
%\addtolength{\textheight}{.2in}
\begin{document}

Dear Editor,\\

We would like to thank the reviewers for their constructive comments on our manuscript. We have thoroughly addressed their remarks. For each of them, we provide here a short summary of how we have addressed it. Additionally we provide a version of the manuscript which highlights the differences in the text for ease of comparison.

We hope that with the current modification, you will find our manuscript suitable for publication in the Journal of Chemical Theory and Computation.

@@

\vspace{1em}

{\bf Editorial Comments}

\begin{formal}
Please include an accessed by date for any reference citing a URL.
\end{formal}

We have corrected the URL references to include an ``Accessed" field.

@@

\clearpage

{\bf Reviewer 1}

\begin{formal}
The article ``The MRChem multiresolution analysis code for molecular electronic structure calculations: performance and scaling properties" by Wind and coauthors describes implementation and parallel performance of the MRChem program package. The paper is important and should definitely be published, but the presentation is not very concise and can be improved.
\end{formal}

We thank the reviewer for their assessment of our work.

\begin{formal}
To me the article has too many implementation details, that distract from the core message of the article, which would be performance and scaling with system size.
\end{formal}

We decreased the level of detail in the paper by moving Sections 2.1 (\emph{The basic operations in the SCF iterations}) and 3 (\emph{Function representation with Multiwavelets}) to the Supporting Information. They are now Sections SI 2 and SI 1, respectively, in the accompanying PDF.

@@

\begin{formal}
General remarks
 - Section 3 can be omitted entirely, in my opinion it is sufficient to state that the space is adaptively partitioned into 3d nodes/intervals. If necessary this chapter can be shortened and incorporated into Section 4
 \end{formal}

 Section 3 has been moved to the Supporting Information as Section SI 1.

@@

 \begin{formal}
 - I also feel that Sec 2.1 could be better included in Sec. 4
 \end{formal}

Section 2.1 has been moved to the Supporting Information as Section SI 2.

\begin{formal}
 - The level of detail is unnecessarily high in Section 4, e.g. explaining the ``compressed" and ``reconstructed" tree states
 \end{formal}

 We have left mentions of the compressed and reconstructed representations as we believe this is needed to understand how screening works in the MRA settings.

 \begin{formal}
 - The authors should always make clear what a ``node" is: an MRA node (3d interval) or a compute node (possible also an MPI rank?). Also, they write about storing data ``together" or ``separately" which I assume refers to storing them on the same or different compute nodes? The same is true for statements that data are ``locally accessible". This should be clarified.
 \end{formal}

 We stated more explicitly when a ``node" is meant as a node in the MRA tree representation and when it is instead meant as a compute node. 
 We clarified the meaning of storing data ``together" and ``separately", as well as explaining that ``locally accessible" refers to data local to the given compute node.

 \begin{formal}
 - Sec 4.2: the naming ``MW nodes in real space" confuses me, but I cannot come up with a better suggestion.
 \end{formal}

 The term ``MW node" is commonly used in the MW literature to refer to each cube in the grid where the scaling and wavelet functions are supported. We have at least tried to be as consistent as possible in the text specifying ``MW node" unless it was obvious (e.g repetition in the same sentence).

\begin{formal}
Minor remarks:
 - Sec 2.1: What is the difference between ``Product of two MW functions" and ``Application of a multiplicative operator"?
 \end{formal}

 There is, technically, no difference, as both a multiplicative operator and MW function(s) are represented as trees. There might be properties of the operator that allow for its more efficient application, thus it is convenient to keep the treatment of the two cases separate.

@@

 \begin{formal}
 - Sec 2.1.1: derivatives are defined in a weak sense, but they are numerically not well-behaved.
 \end{formal}

 This section has now been moved to the SI. The issue of derivatives in a Multiwavelet representation has been investigated both formally and numerically. The latest publication, referenced in Section 2, is by Anderson \emph{et. al.} (DOI: 10.1016/j.jcpx.2019.100033). As written in the text, as long as one avoids ``repeated iterations" (\emph{e.g.} time propagation) their application does not pose a fundamental problem. By applying the integral Helmholtz operator at each iteration the numerical artifacts caused by derivatives, are effectively removed.

@@

 \begin{formal}
 - Sec 2.1.2 and 2.1.3: summing and applying does not commute: is there any numerical difference between both approaches?
 \end{formal}

% The two approaches can lead to different numerical results, since floating-point arithmetic is not associative. This is especially true in a parallel implementation, like ours. MRChem offers an option (\texttt{numerically\_exact}) that forces the use of MPI algorithms giving results invariant with respect to the number of compute nodes.

 The two approaches can lead to different results, since the adaptivity may truncate the intermediates. The differences in the final results should however always be within the required precision.

@@

 \begin{formal}
 - Eq. 17: $\sum_p$ instead of $\sum_k$
 \end{formal}

This has been fixed.

\begin{formal}
Suggestions:
 - The authors state correctly that data sparsity comes with MRA, even if orbitals have non-vanishing contributions from all regions. This is an important statement and has implications for both performance and precision and could be placed more prominently.
\end{formal}

A sentence in the abstract and one in the introduction have been added to underline the implication for both performance and precision of using a MW representation.

@@

\begin{formal}
 - The authors might mention that many first-order molecular properties, including gradients, can be computed at little extra cost via the Hellmann-Feynman theorem.
\end{formal}

A sentence emphasizing the possibility to exploit formal results such as the Hellmann--Feynman theorem has been added at the end of Section 2

@@

\begin{formal}
 - It would be interesting to see scaling properties with respect to MPI ranks and compute nodes for a given molecule/precision.
\end{formal}

We added Section 4.3 (\emph{Scaling with number of compute nodes}) with an analysis of scaling with respect to the number of compute nodes for the case of valinomycine at MW4 level.

### 2022-11-06 SI.tex
Updates from Overleaf

\documentclass[journal=jctcce, manuscript=article]{achemso}

@@

\usepackage[table]{xcolor}
\usepackage[version=4]{mhchem}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\graphicspath{{img/}}
 \DeclareGraphicsExtensions{.pdf} %png for low res. pdf for high res. I also have svg as well but other packages needed
\usepackage{caption}
\usepackage{subcaption}
 \DeclareCaptionSubType*[arabic]{figure}
\usepackage{braket}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{todonotes}
\setlength{\columnsep}{1cm}
\usepackage{booktabs}
\usepackage[nohyperlinks,nolist]{acronym}
\usepackage{siunitx}
\usepackage{mathtools}
% to submit to the arXiv:
% 1. uncomment the following
% 2. clear cached files
% 3. recompile from scratch
%\usepackage[newfloat,finalizecache=true,cachedir=.]{minted}
% then:
% 4. comment the line with the finalizecache option
% 5. uncomment the line with the frozencache option
% 6. recompile (to check that it works)
%\usepackage[newfloat,frozencache=true,cachedir=.]{minted}
\usepackage[newfloat]{minted}
%\captionsetup[listing]{position=top}
\setminted{
  fontsize=\footnotesize,
  xleftmargin=\parindent,
  xrightmargin=\parindent,
  breaklines,
  frame=lines
}
\BeforeBeginEnvironment{minted}{\vspace{-7pt}}
\AfterEndEnvironment{minted}{\vspace{-7pt}}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tablefootnote}
\usepackage{multirow}
%\usepackage{hyperref}

@@

@@

\begin{abstract}
MRChem is a code for molecular electronic structure calculations, based on a multiwavelet adaptive basis representation. We provide a description of our implementation strategy and several benchmark calculations. Systems comprising more than a thousand orbitals are investigated at the \acl{HF} level of theory, with an emphasis on scaling properties. 
With our design, some terms which formally scale quadratically with the system size, in effect have a better scaling because of the implicit screening introduced by the inherent adaptivity of the method. Comparisons with traditional Gaussian-type orbitals based software, show that MRChem can be competitive with respect to performance.
\end{abstract}

\section{Introduction}

\acp{GTO} and more generally \acp{LCAO}\cite{Jensen.2013.10.1002/wcms.1123} are well established as a standard for \emph{ab initio} molecular electronic structure calculations. As their shape is closely related to the electronic structure of atoms, even very small basis sets with only a few functions per \ac{MO} can give reasonable results for describing molecular properties. However, for extended systems the description of each orbital still requires the contributions from the entire basis in order to ensure orthogonality. Without further precautions, even when using localized orbitals, a large proportion of the coefficients will be very small for those systems.

In a \ac{MRA} framework like \acp{MW}, the basis can adapt according to the function described (for an in-depth review of the \ac{MRA} method in the field of Quantum Chemistry, see Ref.~\citenum{Bischoff2019-mr}). The available basis is in principle infinite and complete, and, in practice, it is dynamically adapted to the local shape of the function and the required precision. This can require the real-space basis to comprise millions of elementary functions for each orbital. In this sense, the method starts with a big handicap compared to \ac{LCAO} basis sets. On the other hand, the exponential growth of available computational resources has in recent years enabled \ac{MRA} calculations on systems comprising thousands of atoms.\cite{Ratcliff2020-hk, Harrison2016-lo}

Two main challenges need to be addressed in order to achieve adequate performance: the large number of operations to perform and the large memory footprint. The former is addressed by limiting the numerical operations to those that are strictly necessary to achieve the requested precision: rigorous error control at each elementary operation is the main strength of a \ac{MW} approach, enabling effective screening. The latter is achieved by algorithmic design: beyond a certain system size, not all data can be stored in local (i.e. fast access) memory. 
On modern computers, data access is generally more time consuming than the mathematical operations, especially if the data is not available locally.\footnote{The online interactive chart at \url{https://colin-scott.github.io/personal_website/research/interactive_latency.html} is useful to understand the relative performance of on-chip \emph{vs.} off-chip memory accesses. In particular, it shows how the evolution of memory chips has been lagging behind that of CPUs.} 
The computer implementation must be able to copy large amounts of data efficiently between compute nodes and the algorithm must be devised to reuse the data already available locally, when possible. In this article, we will present the main implementation features of our \ac{MRA} code, MRChem \cite{mrchem}, to tackle those challenges, thus enabling calculations on large systems at the \ac{HF} level.

Beyond the effective thresholding that screens numerically negligible operations, the large computational cost is addressed by parallelization either at the orbital level or through real space partitioning. This dual strategy allows to minimize the relative cost of communication and the overall computational costs. Further, the most intensive mathematical operations are expressed as matrix multiplications, which allows for optimal efficiency.  
Using localized orbitals, the adaptivity of the \ac{MW} description will significantly reduce the computational effort required to compute the terms involving remote orbitals. This opens the way for a method that scales linearly with system size ($N$), where the linearity arises naturally from the methodology, rather than being obtained with \emph{ad hoc} techniques, such as fast-multipole methods \cite{GREENGARD1987325} for Fock matrix construction, or purification of the density matrix. \cite{Helgaker2000-yb}

The large memory requirement is addressed by storing the data in a distributed ``memory bank", where orbitals can be sent and received independently by any CPU. The total size of the memory bank is then only limited by the overall memory available on the entire computer cluster.
Benchmark calculations at the \ac{HF} level show that MRChem is able to describe systems with thousands of electrons. The implementation exhibits near-linear scaling properties and it can also be competitive with state-of-the-art electronic structure software based on \ac{LCAO} methods.

\section{Solving the Hartree--Fock equations with multiwavelets}

We consider the \ac{SCF} equations of the \ac{HF} method:
\begin{equation}\label{eq:differential-SCF}
  \left(T + V \right) \orbital_i = \sum_{j}^{N_{\mathrm{occ}}} F_{ij} \orbital_j,
\end{equation}
where the Fock matrix elements $F_{ij}$ are obtained as
\begin{equation}\label{eq:fock-matrix-definition}
  F_{ij} = \Braket{ {\orbital_i}|{T+V}|{\orbital_j}}.
\end{equation}
In the above equations, $T = -\frac{1}{2}\nabla^2$ is the kinetic
energy operator, and the potential $V$ contains the nuclear-electron
attraction $V_{\mathrm{nuc}}$, the Coulomb repulsion $J$, and the exact exchange $K$. 
To solve the \ac{SCF} equations within a \ac{MW} framework, we rewrite the differential equation \eqref{eq:differential-SCF} as an integral equation:
\begin{equation}\label{eq:integral-SCF}
\orbital_i = -2 H^{\mu_i}\star\left[V\orbital_i - \sum_{j \ne i}^{N_{\mathrm{occ}}} F_{ij} \orbital_j\right],
\end{equation}
where $\star$ stands for the convolution product.
This form is obtained by recalling that the shifted kinetic energy operator $T-\epsilon$ has a closed-form Green's function:\cite{Kalos1962-ok,Beylkin2005-kg}
\begin{equation}\label{eq:helmholtz-green}
    (T-\epsilon)^{-1} = 2 H^{\mu}, \quad H^{\mu} = \frac{e^{-\mu r}}{r}, \quad \mu = \sqrt{-2\epsilon},
\end{equation}
By setting $\epsilon_i = F_{ii}$ and $\mu_i = \sqrt{-2F_{ii}}$, the diagonal term is excluded from the summation in Eq.~\eqref{eq:integral-SCF}.

A \ac{MW} representation does not have a fixed basis. It is therefore not possible to express functions and operators as vectors and matrices of a predefined size, and the virtual space is to be regarded as infinite. Still, each function has a finite, truncated representation defined by the chosen precision threshold, but this representation will in general be different for different functions. It is therefore necessary to develop working equations which allow the direct optimization of orbitals. On the other hand, only occupied orbitals are constructed and the coupling through the Fock matrix in Eqs. \eqref{eq:differential-SCF} and \eqref{eq:integral-SCF} is therefore limited in size.

Differential operators such as the Laplacian pose a fundamental problem for function representations which are not continuous, and a naive implementation leads to numerical instabilities. Robust approximations are nowadays available\cite{Anderson2019-bx}, but for repeated iterations avoiding the use of the Laplacian operator is still an advantage. This is the main reason for using the integral form in Eq.~\eqref{eq:helmholtz-green} rather than the differential form.

Additionally, it can be shown that iterating on the integral equation corresponds to a preconditioned steepest descent. Practical applications show that this approach is comparable in efficiency (as measured in the rate of convergence of the \ac{SCF} iterations) to more traditional methods.\cite{Harrison2004-fs}

\subsection{The basic operations in the SCF iterations}

@@

\subsubsection{The kinetic operator $T$}
\label{sec:compute-T}
The application of the kinetic energy operator on a \ac{MW} function is formally not well defined due to the discontinuities in the basis.\cite{Anderson2019-bx} However, when adopting an integral reformulation of the \ac{SCF} equations, the kinetic energy operator is required only in the  expectation value expressions as in Eq.~\eqref{eq:fock-matrix-definition}, and as shown by \citeauthor{Anderson2019-bx}~\cite{Anderson2019-bx} the original definition of the first-order derivative by \citeauthor{Alpert:2002cx}~\cite{Alpert:2002cx} is able to deliver sufficient precision in this specific case. On the basis of this consideration, and for computational efficiency, the matrix elements of the Laplacian operator are computed as inner products of orbital gradients:
\begin{equation}
  \label{eq:kinetic-matrix}
  T_{ij} = 
  -\frac{1}{2} 
  \Braket{\orbital_i | \nabla^2 \orbital_j} 
  = 
  \frac{1}{2} 
  \Braket{\nabla \orbital_i | \nabla \orbital_j}.
\end{equation}

@@

\subsubsection{The nuclear potential $V_{\mathrm{nuc}}$}
\label{sec:compute-V}
The nuclear potential consists of a sum of contributions from the different nuclei. As such, applying it to a given orbital can be performed in two ways.

@@

The first approach requires $N_{\mathrm{atoms}}\times N_{\mathrm{occ}}$ function multiplications, followed by a summation of $N_{\mathrm{atoms}}$ terms. It also needs memory for $N_{\mathrm{atoms}}\times N_{\mathrm{occ}}$ \ac{MW} functions. However, since these objects are relatively localized, their memory footprint is modest and can, in addition, be amortized by distributing objects across compute nodes. Furthermore, the function-function products can also be distributed across nodes and screened aggressively.

@@

\subsubsection{The Coulomb potential $J$}\label{sec:compute-J}
The Coulomb interaction can be expressed as:
\begin{equation}
    \begin{aligned}
  J(\rvec) &= \sum_i \int \frac{\orbital^*_i(\rvec')\orbital_i(\rvec')}{|\rvec-\rvec'|} \, \mathrm{d}\rvec' =
  \sum_i  \left[
  \frac{1}{|\rvec-\rvec'|} \star 
  \orbital^*_i\orbital_i
  \right] =
  \frac{1}{|\rvec-\rvec'|} \star 
  \left[\sum_i 
  \orbital^*_i\orbital_i
  \right]
    \end{aligned}
\end{equation}
where, as indicated by the last step, we can choose the order of the operations when constructing the Coulomb potential: either summing the orbital densities first and compute the potential of the total density (\textbf{sum-then-apply}), or, first computing the Coulomb potential of each individual orbital density and then sum the potentials (\textbf{apply-then-sum}). The electrostatic potential is in general a much smoother function than its corresponding density, which contains cusps at every nuclear site. It is thus desirable to avoid the explicit construction of the global density and instead gradually build a global potential by partial applications of the Poisson operator to subsets of orbital densities. This operation can also easily be distributed in parallel by assigning different subsets to different processes.

@@

\subsubsection{The exchange potential $K$}\label{sec:compute-K}
The construction and application of the exchange potential is the most computationally demanding part, because the exchange operator cannot be written as a simple multiplicative potential. A Coulomb potential must be constructed and applied for each orbital pair:

@@

\subsubsection{Construction of the right-hand side and Helmholtz operator application}\label{sec:rhs}
The central step in the SCF iteration is the application of the Helmholtz operator $H^{\mu}$ defined in Eq.~\eqref{eq:helmholtz-green} to the square bracket on the right-hand side in Eq.~\eqref{eq:integral-SCF}. The Helmholtz kernel depends on the orbital energy ($\mu_i = \sqrt{-2F_{ii}}$). As such, one needs to construct as many operators as there are occupied orbitals. Given the Fock matrix, each term within the square bracket in Eq.~\eqref{eq:integral-SCF} can be assembled with $O(N)$ formal operation count per orbital $\orbital_i$. Thus, the formation of the whole right-hand side for the entire set of orbitals formally scales as $O(N^{2})$. However, $\sum_{j \ne i}^{N_{\mathrm{occ}}} F_{ij} \orbital_j $ is a linear combination of orbitals and, as shown in Section~\ref{sec:localvsorb}, it can be performed efficiently with $O(N)$ operation count. Also the application of the exchange operator can be done essentially in $O(N)$ operations as shown in Sections \ref{Xscreen} and \ref{sec:screen}.

@@

\section{Implementation details and parallelization strategy}

\ac{MW} calculations can be computationally demanding. Both the total amount of elementary operations, the large amount of memory required and the capacity of the network are important bottlenecks.

In practice, a supercomputer is required for calculations on large systems.
For example the representation of one orbital can demand between 10 and 500MB of data, depending on the kind of orbital and the precision requested (see Section \ref{sizes} for details). Moreover, several sets of functions are necessary in a single \ac{SCF} iteration (orbitals at the previous iterations, right-hand side of equations etc.). Large systems will eventually require more memory than is locally available on a single compute node in a cluster, and distributed data storage becomes mandatory. At the same time, the \ac{SCF} equations will require pairs of orbitals, i.e. all the data must at some point be accessible locally.

For an efficient solution scheme on a cluster architecture, the problem needs to be partitioned into smaller portions that can be computed with as little transfer of data as possible between the different nodes of the cluster.

\subsection{Localized versus canonical orbitals}

Although localized and canonical orbitals represent the same $N$-electron wave function, the numerical operations required to solve the equations will have different scaling properties in a canonical or localized representation.
Because of the inherent adaptivity of the \ac{MW} approach, multiplying two orbitals, or applying an operator onto an orbital can be done more efficiently if the orbitals are localized in space. Even if the orbitals have non-negligible values in a large portion of space, the adaptivity properties will make sure that the calculations in the parts where the orbitals are small are done at a coarser level of accuracy, and therefore faster than in the regions with larger values.

We use the Foster--Boys localization criterion.\cite{Foster1960} The unitary, localizing orbital rotation is computed by minimizing the sum of the occupied orbitals' second central moment.\cite{Hoyvik2016-cw} See also sections \ref{sec:localvsorb} and \ref{sec:N3terms} for more details on the practical implementation.

\subsection{Local (real-space) versus orbital representation}
\label{sec:localvsorb}
%NB: easy to confuse real space local and localized orbitals!

\begin{figure}
\centering
\includegraphics[width=1.\textwidth]{checkerboard.png}
\caption{\label{fig01} Real-space vs. orbital representation. A dark cell represent a node occupied by the corresponding orbital. In an orbital representation (rows) the data from each row is stored and retrieved together. In a local representation (columns), the data from each column is stored and used together. The most convenient representation depends on the operation at hand and it is therefore best to be able to switch from one to the other.}
\end{figure}

Two main strategies can be adopted to distribute the memory storage requirements for orbitals throughout the cluster. The first strategy follows \emph{verbatim} the mathematical expressions: the \ac{HF} equations are written in terms of orbitals, and therefore it is natural to keep the data representing each orbital together. The second strategy, is based on a real-space partitioning: contributions from all the orbitals in a given portion of real space are stored together (see Figure~\ref{fig01}). Both strategies have advantages and disadvantages, and in practice it is convenient to switch between them depending on the given operation.

Taking as an example the computation of the Fock matrix elements in Eq.~\eqref{eq:fock-matrix-definition},  we will show why it is advantageous to shift to a real-space partitioning approach.\footnote{A similar development has happened for the \ac{LCAO} methods: they first were based on \acp{MO}, but when the focus shifted towards large systems, it was soon realized that efficient implementations demanded a design based on using \acp{AO} directly.\cite{Helgaker2000-yb}}
Operator matrix elements are evaluated as

\begin{equation}
  V_{ij} = \braket{\orbital_{i} | {\hat V} |\orbital_j}  = \braket{\orbital_i | {\hat V} \orbital_j} 
\end{equation}
where $\orbital_i$ and $\orbital_j$ are occupied orbitals and ${\hat V}$ is any one-electron operator, such as the nuclear potential operator. If the application of the operator ($\ket{{\hat V} \orbital_j},\,\,\forall j$) is considered as a $O(N_{\mathrm{occ}})$
%not if the operator grows with N. Not that simple!
operation, computing the entire matrix scales formally as $O(N_{\mathrm{occ}}^2)$.

Both the $\orbital_i$
and the result of the operator application $\tau_j = \hat{V} \orbital_j$
are represented either through the \emph{compressed} or \emph{reconstructed} representation, as described in Section~\ref{sec:func_rep}. For instance the compressed representations for $\orbital_i$ and $\tau_i$ would yield respectively
\begin{equation}\label{eq:compressed_orb}
  \orbital_i = \sum_{p} c^i_p \scaling_p + \sum_{np} d^i_{np} \wavelet_{np} \quad\text{and}\quad \tau_j = {\hat V} \orbital_j = \sum_{p} t^j_p \scaling_p + \sum_{np} w^j_{np} \wavelet_{np}.
\end{equation}
Where $n$ represents the composite index $sl$ and runs over all the included \ac{MW} nodes.

When representing functions in 3 dimensions,
using a tensor-product basis and localized orbitals, both $p$ and the number of \ac{MW} nodes, are typically of size $\sim 10^4$.

To perform a scalar product, it is sufficient to recall that the scaling/wavelet basis is fully orthonormal:
\begin{equation}\label{eq:wavelet_othrogonality}
  \braket{\scaling_{p} | \scaling_{p'}}  = \delta_{pp'}, \quad   \braket{\wavelet_{np} | \wavelet_{n'p'}}  = \delta_{nn'}\delta_{pp'}, \quad   \braket{\scaling_{p} | \wavelet_{n'p'}}  = 0  
\end{equation}

which then yields the following result for an operator matrix element between two orbitals:
\begin{equation}
  \braket{\orbital_i|{\hat V} \orbital_j} = \sum_{k} c^i_{p} t^j_{p} + \sum_{np} d^i_{np} w^j_{np}
\end{equation}
where now the sum over the nodes $n$ only extends to those nodes which are present for both functions.

Ignoring for simplicity the scaling contribution which has negligible impact on the performance, this operation is equivalent to a matrix multiplication $D \cdot W^T$ with $D_{ia} = d^i_{np}$ and $W_{ja} = w^j_{np}$, and $a$ is the compound-index $a=np=nsl$. However, for large systems it is not possible to hold the full matrices in local memory.
The \ac{MO}-like approach is then to sum over all compound indices $a$ for each pair $i,j$ independently. In this way each pair of coefficients $d^i_{np}$ and $w^k_{np}$ is used only once after each memory fetch. On modern computers, memory access is much more time-consuming than the mathematical operations themselves, up to a factor 100 for most CPU-based systems. A faster implementation is to consider a specific \ac{MW} node $n$ and perform the product for all $i$, $j$ and $p$ values within this node:
\begin{equation}\label{equ:sum}
  \braket{\orbital_i | {\hat V} |\orbital_j}_{n} 
  = \sum_{p} \left(d^i_{p} w^j_{p}\right)_{n}, \quad\forall i,j,
\end{equation}
this results in a series of independent matrix multiplications, thus fully utilizing the capabilities of the computer.

To perform the operation in this way, the storage of the orbitals in memory must be adapted. Instead of storing each orbital separately (orbital or MO representation), the values of all the orbitals at specific positions in space (\emph{i.e.}~a \ac{MW} node) must be easily accessible (local representation). Switching from an orbital to a local representation is a $O(N)$ operation, where $N$ measures the system size, typically the number of orbitals.
If all $i$ and $j$ are included, the number of operations still scales formally as the square of the number of orbitals. However for a given node $n$, the products only needs to be explicitly performed if both functions (indices $i$ and $j$) have a non-negligible contribution to the node $n$. When orbitals are localized, the number of such contributions will eventually scale linearly with the system size.

Linear combinations of orbitals can be addressed similarly. Let us consider the linear combination
\begin{equation}
  \tilde{\orbital}_i = \sum_{j} A_{ij} \orbital_j
\end{equation}
where $A_{ij}$ is the coefficient matrix of the transformation.
In a linear transformation, each \ac{MW} node can be transformed independently for each node $n$, to obtain the corresponding coefficients for $\tilde{\orbital}_i$:
\begin{equation}
  w^i_{np} = \sum_{j} A_{ij} d^j_{np}
\end{equation}
Here also, the number of localized orbitals contributing to a specific \ac{MW} node $n$ grows as $O(1)$ with system size. 
When constructing $\tilde{\orbital}_i$, the grid is successively refined, so that only values corresponding to \ac{MW} nodes with wavelet norms above the required threshold are actually computed.
Moreover, due to linearity this operation is carried out both for scaling and wavelet coefficients, thus avoiding the need to perform any \ac{MW} transform.

\subsection{Adaptivity and precision thresholding}\label{sec:adaptivity-screening}

The most powerful idea of \ac{MRA} is to limit the precision of intermediate operations to what is strictly necessary to achieve the requested precision in the final result. This guiding principle can be exploited at several levels:

\begin{enumerate}
    \item Elementary mathematical operations on a function in a multiwavelet representation will automatically refine the local basis (\emph{i.e.}~grid) of the resulting function, but only until the required precision is achieved. The computational effort required will vary according to the position and shape of the functions or operators. For example, the product of two functions which are localized in different parts of space, will normally be done faster than the square of each of them.
    \item Only those parts of functions with absolute value (norm of the \ac{MW} node) above a defined threshold will be retrieved before even processing them; this is used \emph{e.g.}~in the scheme presented in the previous section.
    \item A \ac{SCF} calculation can be initiated at low precision and the final result can be used as starting guess for a calculation with higher precision. In contrast to finite \ac{AO} basis sets, which are usually not subsets of each other, a multiwavelet function computed at a given precision, can be directly used as a valid representation at a different precision.
\end{enumerate}

\subsection{Exchange screening}
\label{Xscreen}

The exchange terms require the evaluation of

\begin{equation}
\label{equ:ex}
  K_i \orbital_j = \orbital_i(\rvec) 
  \left[
  \int \frac{\orbital^*_i(\rvec')\orbital_j(\rvec')}{|\rvec-\rvec'|} \, \mathrm{d}\rvec'
  \right]
\end{equation}
 for all $\frac{N_{\mathrm{occ}}(N_{\mathrm{occ}}-1)}{2}$ orbital pairs $i,j$.

>>>>

\documentclass[journal=jctcce, manuscript=suppinfo]{achemso}

@@

\usepackage[table]{xcolor}
\usepackage[version=4]{mhchem}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\graphicspath{{img/}}
 \DeclareGraphicsExtensions{.pdf} %png for low res. pdf for high res. I also have svg as well but other packages needed
\usepackage{caption}
\usepackage{subcaption}
 \DeclareCaptionSubType*[arabic]{figure}
\usepackage{braket}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{todonotes}
\setlength{\columnsep}{1cm}
\usepackage{booktabs}
\usepackage[nohyperlinks,nolist]{acronym}
\usepackage{siunitx}
\usepackage{mathtools}
% to submit to the arXiv:
% 1. uncomment the following
% 2. clear cached files
% 3. recompile from scratch
%\usepackage[newfloat,finalizecache=true,cachedir=.]{minted}
% then:
% 4. comment the line with the finalizecache option
% 5. uncomment the line with the frozencache option
% 6. recompile (to check that it works)
%\usepackage[newfloat,frozencache=true,cachedir=.]{minted}
\usepackage[newfloat]{minted}
%\captionsetup[listing]{position=top}
\setminted{
  fontsize=\footnotesize,
  xleftmargin=\parindent,
  xrightmargin=\parindent,
  breaklines,
  frame=lines
}
\BeforeBeginEnvironment{minted}{\vspace{-7pt}}
\AfterEndEnvironment{minted}{\vspace{-7pt}}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tablefootnote}
\usepackage{multirow}
%\usepackage{hyperref}
\usepackage{xr}

\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}
\myexternaldocument{main}

@@

\renewcommand{\theequation}{SI \arabic{equation}}
\renewcommand{\thesection}{SI \arabic{section}}

@@

A note on this document: equations and sections referenced in the Supporting Information refer to the main manuscript, unless the reference number contains the SI prefix.

\section{The basic operations in the SCF iterations}

@@

\subsection{The kinetic operator $T$}
\label{sec:compute-T}
The application of the kinetic energy operator on a \ac{MW} function is formally not well defined due to the discontinuities in the basis.\cite{Anderson2019-bx} However, when adopting an integral reformulation of the \ac{SCF} equations, the kinetic energy operator is required only in the  expectation value expressions as in Eq.~\eqref{eq:fock-matrix-definition}, and as shown by \citeauthor{Anderson2019-bx}~\cite{Anderson2019-bx} the original definition of the first-order derivative by \citeauthor{Alpert:2002cx}~\cite{Alpert:2002cx} is able to deliver sufficient precision in this specific case. On the basis of this consideration, and for computational efficiency, the matrix elements of the Laplacian operator are computed as inner products of orbital gradients:
\begin{equation}
  \label{eq:kinetic-matrix}
  T_{ij} = 
  -\frac{1}{2} 
  \Braket{\orbital_i | \nabla^2 \orbital_j} 
  = 
  \frac{1}{2} 
  \Braket{\nabla \orbital_i | \nabla \orbital_j}.
\end{equation}

@@

\subsection{The nuclear potential $V_{\mathrm{nuc}}$}
\label{sec:compute-V}
The nuclear potential consists of a sum of contributions from the different nuclei. As such, applying it to a given orbital can be performed in two ways.

@@

The first approach requires $N_{\mathrm{atoms}}\times N_{\mathrm{occ}}$ function multiplications, followed by a summation of $N_{\mathrm{atoms}}$ terms. It also needs memory for $N_{\mathrm{atoms}}\times N_{\mathrm{occ}}$ \ac{MW} functions. However, since these objects are relatively localized, their memory footprint is modest and can, in addition, be amortized by distributing objects across parallel processes. Furthermore, the function-function products can also be distributed across nodes and screened aggressively.

@@

\subsection{The Coulomb potential $J$}\label{sec:compute-J}
The Coulomb interaction can be expressed as:
\begin{equation}
    \begin{aligned}
  J(\rvec) &= \sum_i \int \frac{\orbital^*_i(\rvec')\orbital_i(\rvec')}{|\rvec-\rvec'|} \, \mathrm{d}\rvec' =
  \sum_i  \left[
  \frac{1}{|\rvec-\rvec'|} \star 
  \orbital^*_i\orbital_i
  \right] =
  \frac{1}{|\rvec-\rvec'|} \star 
  \left[\sum_i 
  \orbital^*_i\orbital_i
  \right]
    \end{aligned}
\end{equation}
where, as indicated by the last step, we can choose the order of the operations when constructing the Coulomb potential: either summing the orbital densities first and compute the potential of the total density (\textbf{sum-then-apply}), or, first computing the Coulomb potential of each individual orbital density and then sum the potentials (\textbf{apply-then-sum}). The electrostatic potential is in general a much smoother function than its corresponding density, which contains cusps at every nuclear site. It is thus desirable to avoid the explicit construction of the global density and instead gradually build a global potential by partial applications of the Poisson operator to subsets of orbital densities. This operation can also easily be distributed in parallel by assigning different subsets to different processes.

@@

\subsection{The exchange potential $K$}\label{sec:compute-K}
The construction and application of the exchange potential is the most computationally demanding part, because the exchange operator cannot be written as a simple multiplicative potential. A Coulomb potential must be constructed and applied for each orbital pair:

@@

\subsection{Construction of the right-hand side and Helmholtz operator application}\label{sec:rhs}
The central step in the SCF iteration is the application of the Helmholtz operator $H^{\mu}$ defined in Eq.~\eqref{eq:helmholtz-green} to the square bracket on the right-hand side in Eq.~\eqref{eq:integral-SCF}. The Helmholtz kernel depends on the orbital energy ($\mu_i = \sqrt{-2F_{ii}}$). As such, one needs to construct as many operators as there are occupied orbitals. Given the Fock matrix, each term within the square bracket in Eq.~\eqref{eq:integral-SCF} can be assembled with $O(N)$ formal operation count per orbital $\orbital_i$. Thus, the formation of the whole right-hand side for the entire set of orbitals formally scales as $O(N^{2})$. However, $\sum_{j \ne i}^{N_{\mathrm{occ}}} F_{ij} \orbital_j $ is a linear combination of orbitals and, as shown in Section~\ref{sec:localvsorb}, it can be performed efficiently with $O(N)$ operation count. Also the application of the exchange operator can be done essentially in $O(N)$ operations as shown in Sections \ref{Xscreen} and \ref{sec:screen}.

@@
